---
title: "Project Document"
format: pdf
editor: visual
---

# Part I: Exploratory Data Analysis

\newpage

## About the Data

Our data was obtained via UCI's Machine Learning Repository. The data is a multivariate set designed to explore student performance tied to various predictors. The data is split into two sets: Mathematics (student-mat.csv) and Portuguese (student-por.csv). These are the two subjects where records of the students performance (our outcome *y*) is recorded. Predictor variables include a range of demographic, social, health, and school related attributes.

The data was utilized by a paper published in 2008 titled ["Using data mining to predict secondary school performance"](https://www.semanticscholar.org/paper/Using-data-mining-to-predict-secondary-school-Cortez-Silva/61d468d5254730bbecf822c6b60d7d6595d9889c). The study's goal was to use BI/DM techniques to build a model that accurately predicted student performance given predictor variables that provided the best accuracy. Below, we will conduct an EDA exploring and cleaning this data set prior to conducting a replication of their study while critiquing their process and adding/removing anything we deem necessary to result in the best models for our given data and prior proposed research goal.

### Loading our Libraries

```{r}
library(tidyverse)
```

### Loading our Data

```{r}
# loading in both of our data sets, we will title them by their 
# subject
math <- read.csv("student-mat.csv", sep =  ";")
portuguese <- read.csv("student-por.csv", sep = ";") # our data was seperated by semi colons 
# instead of the traditional comma
```

\newpage

## Understanding the Data Structure

Prior to inspecting and cleaning our data, it is important we fully encapsulate what each column, row, and value mean.

```{r}
head(math)
head(portuguese)
```

Both of our data sets are structured the same with the same column and row variables as well as structured values. This will help make implementing any cleaning and modeling simpler.

### Variables & Values

Referring back to the original paper, there are 33 columns of interest. Those variables are listed below alongside their values, with explanation and clarification as needed, below:

**For a visual example, I have printed a random row to show how the values are presented as they are explained below**

```{r}
math[3, ]
```

`school` - Binary values of either `GP` (Gabriel Pereira) or `MS` (Mousinho da Silveira) of which school a student attended.

`sex` - Binary values of either `F` (female) or `M` (male) regarding a students sex.

`age` - Numeric value of a students age from 15 - 22.

`address` - Binary values of either `U` (urban) or `R` (rural) regarding a students home address.

`famsize` - Binary values of either `LE3` (less than or equal to 3 family members) or `GT3` (greater than 3 family members).

`Pstatus` - Binary values of either `T` (parents are living together) or `A` (parents are living apart) for parents living status.

`Medu` - Leveled integer value of range 0-4 with 0 reflecting no education or below primary completion, 1 reflecting completion of primary education (up to 4th grade), 2 reflecting completion of 5-9th grade education, 3 reflecting completion of secondary education, and 4 reflecting higher education (college degree or higher) of a students' mother's education.

`Fedu` - Leveled integer value of range 0-4 with 0 reflecting no education or below primary completion, 1 reflecting completion of primary education (up to 4th grade), 2 reflecting completion of 5-9th grade education, 3 reflecting completion of secondary education, and 4 reflecting higher education (college degree or higher) of a students' father's education.

`Mjob` - Nominal values for a students' mother's job classified as `teacher`, `health` (any care related profession), `services` (any administrative or police related field), `at_home` (none), `other` (not stated).

`Fjob` - Nominal values for a students' father's job classified as `teacher`, `health` (any care related profession), `services` (any administrative or police related field), `at_home` (none), `other` (not stated).

`reason` - Nominal values for a student's reason for school selection as either `home` (close to home), `reputation`, `course` (valued courses provided), `other` (reason not stated).

`guardian` - Nominal values for who the primary caregiver of the student is as either `mother`, `father`, or `other`. Reason for why both parents cannot be listed is not stated.

`traveltime` - Leveled integer values representing travel time to school on a scale of 1-4, `1` reflecting \<15 minutes, `2` reflecting 15-30 minutes, `3` reflecting 30 minutes to 1 hour, `4` reflecting \>1 hour travel time.

`studytime` - Leveled integer values representing average weekly study time reported by the student on a scale of 1-4, `1` reflecting \<2 hours, `2` reflecting 2-5 hours, `3` reflecting 5-10 hours, `4` reflecting \>10 hours study time.

`failures` - Leveled integer values representing the number of classes a student has failed prior to enrolling in this course with a scale of 1-4, each reflecting the amount of courses failed, 4 being \> or = 4 failed classes.

`schoolsup` - Binary value for either `yes` or `no` student receiving additional educational support outside of the course. Not specified if this is inclusive of in-school tutoring and/or support such as services for language gaps or speech development.

`famsup` - Binary value for either `yes` or `no` student receiving additional family educational support outside of the course (family members assist in helping the student with studying or homework).

`paid` - Binary value for either `yes` or `no` student is paying for additional educational support for the course.

`activities` - Binary value for either `yes` or `no` student is participating in extra-curricular activities.

`nursery` - Binary value for either `yes` or `no` student attended nursery school in the past (equivalent to pre-school education in America).

`higher` - Binary value for either `yes` or `no` student wants to pursue higher education courses in the future.

`internet` - Binary value for either `yes` or `no` student has internet access at home.

`romantic` - Binary value for either `yes` or `no` student is currently in a romantic relationship.

`famrel` - Leveled integer values scaled from 1-5 for a students quality of family relationships, `1` being very bad and `5` being excellent.

`freetime` - Leveled integer values scaled from 1-5 for a students free time after school, `1` being very little free time and `5` being lots of free time.

`goout` - Leveled integer values scaled from 1-5 of how often a student goes out with freinds, `1` being not often and `5` being very often.

`Dalc` - Leveled integer values scaled from 1-5 of how often a student consumes alcohol on a weekday, `1` being not often and `5` being very often.

`Walc` - Leveled integer values scaled from 1-5 of how often a student consumes alcohol on a weekend, `1` being not often and `5` being very often.

`health` - Leveled integer values scaled from 1-5 of a students health status, `1` being bad and `5` being very good.

`absences` - Numeric values of the number of day absences the student has from the course so far, i.e. a value of `5` would mean the student has been absent from the class a total of 5 times.

`G1` - Leveled integer values scaled from 0-20 of a students first period grade in the course(period is a trimester in American equivalency).

`G2` - Leveled integer values scaled from 0-20 of a students second period grade in the course.

`G3` - Leveled integer values scaled from 0-20 of a students third period grade in the course.

### Classes & Values

Given the review of our variables and their values, we should expect (was gonna explain what classes i expect them to be and also suggest changing the G1-G3 to numeric values so we can use them unleveled)

```{r}

```

### Cleaning Up

anything extra you may want to add or clean or change

## Correlations & Distribution 

### Means & Distributions

summaries, correlations, plots, anything prior to statistical analysis

### Correlations & Plots

etc

## Statistical Analysis

### Review of Plan to fit Models

talk about models, which they used in the paper, hence why we will run certain statistical analysis

### Running Statistical Analysis to Properly Fit Models

statistical analysis necessary to conduct prior to modeling

## Modeling Procedures 

talk about models in paper, models we will replicate, how, and why, etc
